# Simulating Differential Abundance

```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = FALSE)
set.seed(2123)
source("DA_wrapper.R")
```

Before we consider simulating entire microbial communities, with their complex
correlation structures, let's learn simulators for individual taxa. This is
already enough to analyze taxon-level differential abundance approaches.  For
example, at the end of this session, we'll apply a simulator to study the power
and false discovery rate of limma-voom when applied to microbiome data (as
opposed to the bulk RNA-seq data for which it was originally proposed).  Also,
marginal modeling is a first step towards multivariate (community-wide)
modeling, which we'll explore in the next session.

Let's load the necessary packages. Instructions for `scDesigner` and `MIGsim`
can be found in the pre-workshop announcement.  `SummarizedExperiment` is on
Bioconductor, and all other packages are on CRAN.

Let's train a simulator to fit the Atlas dataset. We'll use `bmi_group` as the 
covariate of interest -- we want to see how microbiome composition varies among 
people with different BMI. We found it helpful
to fixed zero inflation across the population (`nu`), so we have set `nu = ~1`.
Finally, since we want to eventually evaluate testing methods that are designed
for count data, we have used the (Z)ero (I)nflated (N)egative (B)inomial
location-shape-scale model.

```{r}
data(atlas1006, package = "microbiome")
atlas <- convertFromPhyloseq(atlas1006) |>
  filter(time == 0) |>
  dplyr::mutate(
    bmi_group = case_when(
      str_detect(bmi_group, "obese") ~ "obese",
      bmi_group == "lean" ~ "lean",
      bmi_group == "overweight" ~ "overweight",
    )
  ) |>
  dplyr::filter(bmi_group %in% c("obese", "lean", "overweight")) |>
  dplyr::mutate(
    bmi_group = factor(bmi_group, levels = c("obese", "overweight", "lean"))
  )

fmla <- list(mu = ~bmi_group, sigma = ~bmi_group, nu = ~1)
```

We'll remove the genera which are never observed in this dataset. 

```{r}
colData(atlas)$log_depth <- log(colSums(assay(atlas)))
zero_var <- apply(assay(atlas), 1, var) == 0
atlas <- atlas[!zero_var, ]
```

Next we will select features that could work with ZINBF() estimation.
```{r}
select_i <- c(1:2)
colData(atlas) <- colData(atlas)[, c("bmi_group", "sample")]

for (i in 3:nrow(atlas)) {
  #print(glue("{i}/{dim(atlas)[1]}"))
  exper_tmp <- atlas[c(select_i, i), ]
  sim <- setup_simulator(exper_tmp, fmla, ~ ZINBF())

  error_warning <- estimate(
    sim,
    control = gamlss.control(n.cyc = 20, mu.step = 0.1)
  ) |>
    tryCatch(warning = \(w) w, error = \(e) e)

  if (is.na(class(error_warning)[2])) {
    select_i <- c(select_i, i)
  }
}
```

```{r, atlas-estimation}
colData(atlas) <- colData(atlas)[, c("bmi_group", "sample")]
atlas_filtered <- atlas[select_i, ]

sim <- setup_simulator(atlas_filtered, fmla, ~ ZINBF()) |>
  estimate(control = gamlss.control(n.cyc = 20, mu.step = 0.1))
```
## Critique

**Exercise**: The block below combines the real and simulated experiments and
visualizes their difference both graphically and quantitatively. With your 
neighbors, discuss how well the simulator approximates the original template.

```{r, contrast-simulators}
combined <- bind_rows(
  real = pivot_experiment(atlas_filtered), # real data
  simulated = pivot_experiment(sample(sim)), # simulated
  .id = "source"
)

top_features <- assay(atlas_filtered) |>
  rowMeans() |>
  sort() |>
  tail(10) |>
  names()

combined |>
  filter(feature %in% top_features) |>
  ggplot() +
  geom_boxplot(
    aes(sqrt(value), reorder(feature, value, median), fill = bmi_group)
  ) +
  facet_grid(. ~ source) +
  labs(x = "Square root transformed counts", y = "", fill = "BMI group") +
  theme(legend.position = "right") +
  scale_fill_manual(values = rev(c("#9491D9", "#3F8C61", "#F24405")))
```
```{r}
kdetest.group <- function(data) {
  x <- data |> filter(source == "real")
  y <- data |> filter(source == "simulated")
  kde.test(x$value, y$value)$pvalue
}

prop.accepted.H0 <- sapply(
  c("lean", "overweight", "obese"),
  \(group) {
    group_filter <- map(
      unique(combined$feature),
      ~ combined |>
        filter(feature == . & bmi_group == group)
    )
    adjpvalue <- sapply(
      seq_along(select_i),
      \(i) tryCatch(kdetest.group(group_filter[[i]]), error = \(e) NA)
    ) |>
      p.adjust(method = "BH")
    mean(adjpvalue > 0.05, na.rm = TRUE)
  }
)
prop.accepted.H0
```

**Solution**: The simulated data are generally similar to the original data. 
However, for some genera with positively skewed distributions and a high number 
of outliers, such as *Prevotella melaninogenica et rel*, the simulation reached
its limits, explaining why the KS two-sample test returned significant
differences between the real and simulated data for some genera. Nonetheless,
the ordering of abundances between the groups typically agrees between the real
and simulated data. The interquartile ranges for each taxon also seem to roughly
match.

## Power Analysis Loop

To run a power analysis, we need to define datasets that have known ground
truth.  Then, we can run any differential abundance methods we want and see how
many of the true associations are recovered (and how many nulls are falsely
rejected). To this end, we'll remove associations from genera based on Wilcoxon 
Rank Sum test.  We'll choose to remove the genera that have a q valuse<0.1
in the original data.This is helpful because, even if we use `bmi_group` in our 
formula, if in reality there is no (or very weak) effect, then even if our 
simulator considers it as a true signal, the difference may be hard to detect. 
Eventually, our package will include functions for modifying these effects 
directly; at this point, though, we can only indirectly modify parameters by 
re-estimating them with new formulas.

```{r, define-null}
wilcox_res <- differential_analysis(atlas_filtered, "wilcox-clr")
nulls <- rownames(wilcox_res[wilcox_res$q_value.bmi_group. > 0.1, ])

sim <- sim |> scDesigner::mutate(any_of(nulls), link = ~1)
sim <- estimate(sim, control = gamlss.control(n.cyc = 20, mu.step = 0.1))
```

Now that we have ground truth associations, we'll evaluate LIMMA-voom, ANCOMBC, 
DESeq2 for differential analysis. We consider sample sizes ranging from 50 to 
1200, and we simulate 20 datasets for each sample size.

```{r, run-loop, eval = TRUE}
config <- expand.grid(
  sample_size = floor(seq(50, 1200, length.out = 5)),
  n_rep = 1:20
) |>
  mutate(run = as.character(row_number()))

results <- list()
for (i in seq_len(nrow(config))) {
  atlas_ <- sample_n_(sim, config$sample_size[i])
  colnames(atlas_) <- glue("Sample-{seq_len(config$sample_size[i])}")
  results[[i]] <- sapply(c("LIMMA_voom", "ANCOMBC", "DESeq2"),
    \(m) {
      differential_analysis(atlas_, m) |>
        da_metrics(nulls)
    },
    simplify = FALSE
  )
  #print(glue("{i}/{nrow(config)}"))
}
```

**Exercise**: Visualize the results. How would you interpret the results of the
power analysis? Based on your earlier critique of the simulator, do you think
the estimated power here is conservative, liberal, or about right?

**Solution**: We'll use the `stat_pointinterval` function from the `ggdist` package to
visualize the range of empirical power estimates across sample sizes. We can see
that the average false discovery proportion is always controlled below 0.1,
though the variance in this proportion can be high. We can also see that
we would have quite good power with $n \geq 625$ samples, but the worst case
scenarios can be quite poor for anything with fewer samples.

```{r, visualize-da-power, eval = TRUE}
power_fdr <- map_dfr(results, ~ bind_rows(., .id = "method"), .id = "run") |>
  dplyr::mutate(sample_size = rep(config$sample_size, each = 6))

ggplot(power_fdr) +
  tidybayes::stat_pointinterval(aes(factor(sample_size), value, group = method, colour = method),
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(aes(yintercept = 0.1, alpha = metric), linetype = 2) +
  facet_wrap(~metric) +
  labs(x = "Sample size", y = "", color = "Method") +
  scale_color_manual(values = wes_palette("Moonrise2", n = 3)) +
  scale_alpha_discrete(guide = "none", range = c(1, 0)) +
  theme(legend.position = "right", legend.box.background = element_rect(colour = "black"))
```

```{r power_plot, include = FALSE, echo = FALSE}
ggsave("~/Downloads/power_plot.png", width = 8, height = 4)
```

```{r}
sessionInfo()
```
